### 2013-07-23 doc/asplos14/abstract.tex
remove caching and clean up some wording
many small fixes

\Grappa is a runtime system for commodity clusters of multicore computers that
presents a massively parallel, single address space abstraction to
applications. \Grappa's purpose is to enable scalable performance of irregular
parallel applications, such as graph processing. Poor data locality,
imbalanced parallel work and complex communication patterns make scaling these
applications difficult.

\Grappa serves both as a C++ user library and as a foundation for higher level
languages. \Grappa tolerates delays to remote memory by multiplexing thousands
of lightweight workers to each processor core; balances load via fine-grained
distributed work-stealing; increases communication throughput by aggregating
smaller data requests into large ones; and provides efficient synchronization
and remote operations. We present a detailed description of the \Grappa system
and performance comparisons on several irregular benchmarks to hand-optimized
MPI code and to the Cray XMT, a custom system used to target the real time
graph analytics market.

>>>>

\Grappa is a runtime system for commodity clusters of multicore computers that
presents a massively parallel, single address space abstraction to
applications. \Grappa's purpose is to enable scalable performance of irregular
parallel applications, such as  branch and bound optimization, SPICE circuit simulation, and graph processing. Poor data locality,
imbalanced parallel work and complex communication patterns make scaling these
applications difficult.

\Grappa serves both as a C++ user library and as a foundation for higher level
languages. \Grappa tolerates delays to remote memory by multiplexing thousands
of lightweight workers to each processor core, balances load via fine-grained
distributed work-stealing, increases communication throughput by aggregating
smaller data requests into large ones, and provides efficient synchronization
and remote operations. We present a detailed description of the \Grappa system
and performance comparisons on several irregular benchmarks to hand-optimized
MPI code and to the Cray XMT, a custom system used to target the real-time
graph-analytics market.

### 2013-07-23 doc/asplos14/communication.tex
remove caching and clean up some wording
many small fixes

\paragraph{Message aggregation} In our experiments the vast majority of upper
layer message requests are smaller than 44 bytes. Our measurements confirm
manufacturers' published data [15]; with 44-byte packets, the available
bisection bandwidth is only a small fraction (3\%) of the peak bisection
bandwidth. As mentioned earlier, commodity networks including InfiniBand
achieves their peak bisection bandwidth only when the packet sizes are
relatively large --- on the order of multiple kilobytes. The reason for this
discrepancy is the combination of overheads associated with handling each
packet (in terms of bytes that form the actual packet, processing time at the
card, multiple round-trips on the PCI Express bus and processing on the CPU
within the driver stack). Consequently, to make the best use of the network,
we must convert small messages into large ones.

\paragraph{Message processing mechanics} Since communication is very frequent
in \Grappa, aggregating and sending messages efficiently is very important. To
achieve that, \Grappa makes careful use of caches, prefetching and lock-free
synchronization operations.

Each processing core of a system node maintains an array of outgoing message lists.  The array size is the number of system nodes in the \Grappa system.   The outgoing message lists and messages are located in a region of memory shared across all cores in a \Grappa node (thus enabling cores to peek at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack), determines the destination system node, and links the buffer into the corresponding linked list.

Each processing core in a given system node is responsible for aggregating and sending the resulting messages from all processors on that node to a set of destination nodes.  Each core periodically executes a task responsible for sending messages.  This task examines the private (to each core) message lists for each destination node it is responsible for managing and if the list is long enough, or a message has waited past a time-out period, all messages to a given destination system node from that source system node are sent.  Aggregating and sending a message involves manipulating a set of shared data-structures (the message lists). This is done using CAS (compare-and-swap) operations to avoid high synchronization costs.  Note that the reason we use a per-processor array of message lists that is only periodically modified across processor cores is we empirically determined this approach was faster (sometimes significantly) than a global per-system node array of message lists.

Each node has a region of memory with send buffers where the final aggregated
messages are built. These buffers are visible to the network card, and
messages can be sent with user-mode operations only. When the worker
responsible for outbound messages to a given system node has received a
sufficient number of message send requests, or a timeout is reached, the
linked list of messages is walked and messages are copied to a send buffer.
This process requires careful prefetching because most of the outbound
messages are \emph{not} in the processor cache at this time (recall that a
core can be aggregating messages originating from other cores in the same
node). Once the send buffer has been formed it is handed off to GASNet for
transfer to the remote system node. RDMA is used if the underlying network
supports it.

There are two useful consequences of forming the send buffer at the time of
message transmission instead of along the way, as individual upper layer
message send requests are received. First, as previously mentioned most of the
messages are not in the cache and prefetching is used to run-ahead in the
linked list of messages in order to avoid cache misses. But once the send
buffer is formed it is in the cache (for the most part). Hence, when it is
handed off to GASNet for transfer across the physical wire, the network card
can pull the message buffer from the processor cache instead of main memory,
which we have found speeds performance. The second consequence of this
decision is that we do not need to pre-allocate buffers for all destination
nodes in the system, as the buffer can be allocated on the fly. Nevertheless
we have found it efficient to build a flow-control like protocol of
outstanding message buffers between pairs of system nodes.

Once the remote system node has received the message buffer, a management task is spawned to manage the unpacking process.  The management task spawns a task on each processor at the receiving system to simultaneously unpack messages destined for that processor.  Upon completion, these unpacking tasks synchronize with the management task.  Once all cores have processed the message buffer the management task sends a reply to the sending system node indicating the successful delivery of the messages.

>>>>

\paragraph{Message aggregation} In our experiments the vast majority of upper
layer message requests are smaller than 44~bytes. Our measurements confirm
manufacturers' published data [15]; with 44-byte packets, the available
bisection bandwidth is only a small fraction (3\%) of the peak bisection
bandwidth. As mentioned earlier, commodity networks including InfiniBand
achieves their peak bisection bandwidth only when the packet sizes are
relatively large -- on the order of multiple kilobytes. The reason for this
discrepancy is the combination of overheads associated with handling each
packet (in terms of bytes that form the actual packet, processing time at the
card, multiple round-trips on the PCI Express bus and processing on the CPU
within the driver stack). Consequently, to make the best use of the network,
we must convert small messages into large ones.

\paragraph{Message processing mechanics} Since communication is very frequent
in \Grappa, aggregating and sending messages efficiently is very important. To
achieve that, \Grappa makes careful use of caches, prefetching, and lock-free
synchronization operations.

Each processing core of a system node maintains an array of outgoing
message lists.  The array size is the number of system cores in the
\Grappa system.  The outgoing message lists and messages are located
in a region of memory shared across all cores in a \Grappa node (thus
enabling cores to peek at each other's message lists). When a task
sends a message, it allocates a buffer (typically on its stack),
determines the destination system node, and links the buffer into the
corresponding linked list.

Each processing core in a given system node is responsible for
aggregating and sending the resulting messages from all processors on
that node to a set of destination nodes.  Each core periodically
executes a task responsible for sending messages.  This task examines
the private (to each core) message lists for each destination node it
is responsible for managing and, if the list is long enough or a
message has waited past a time-out period, all messages to a given
destination system node from that source system node are sent.
Aggregating and sending a message involves manipulating a set of
shared data-structures (the message lists). This is done using CAS
(compare-and-swap) operations to avoid high synchronization costs.
Note that we use a per-processor array of message lists
that is only periodically modified across processor cores because we
empirically determined this approach was faster (sometimes
significantly) than a global per-system node array of message lists.

Each node has a region of memory with send buffers where the final aggregated
messages are built. These buffers are visible to the network card, and
messages can be sent with user-mode operations only. When the worker
responsible for outbound messages to a given system node has received a
sufficient number of message send requests or a timeout is reached, the
linked list of messages is walked and messages are copied to a send buffer.
This process requires careful prefetching because most of the outbound
messages are \emph{not\/} in the processor cache at this time (recall that a
core can be aggregating messages originating from other cores in the same
node). Once the send buffer has been formed, it is handed off to GASNet for
transfer to the remote system node. RDMA is used if the underlying network
supports it.

There are two useful consequences of forming the send buffer at the time of
message transmission instead of along the way as individual upper layer
message send requests are received. First, as previously mentioned, most of the
messages are not in the cache and prefetching is used to run ahead in the
linked list of messages in order to avoid cache misses. But once the send
buffer is formed, it is in the cache (for the most part). Hence, when it is
handed off to GASNet for transfer across the physical wire, the network card
can pull the message buffer from the processor cache instead of main memory,
which we have found speeds performance. The second consequence of this
decision is that we do not need to pre-allocate buffers for all destination
nodes in the system, as the buffer can be allocated on the fly. Nevertheless
we have found it efficient to build a flow-control-like protocol of
outstanding message buffers between pairs of system nodes.

Once the remote system node has received the message buffer, a
management task is spawned to manage the unpacking process.  The
management task spawns a task on each processor at the receiving
system to simultaneously unpack messages destined for that processor.
Upon completion, these unpacking tasks synchronize with the management
task.  Once all cores have processed the message buffer, the management
task sends a reply to the sending system node indicating the
successful delivery of the messages.

### 2013-07-23 doc/asplos14/evaluation.tex
many small fixes

Our questions to answer with experimentation are:
The goal of our evaluation is to show that the core pieces of the \Grappa
runtime system, namely our tasking system and the global memory/communication
layer work as expected and together are able to efficiently run irregular
applications. We evaluate \Grappa in three basic steps:

@@

\item We present results that show that \Grappa can
    support large amounts of concurrency: sufficient for remote memory
    access and aggregation. The communication layer is able to
    achievesustain a very high rate of global memory operations. We also show the performance of a
graph kernel that stresses communication and concurrency together.
\TODO{compare to key-value stores?}

@@

Figure~\ref{fig:context-switch-exp} shows the average context switch
time as the number of workers grow. At our standard operating point
($\approx$1K workers), context switch time is on the order of 50
ns. As we add workers, the time increases slowly, but levels off: we
also ran with 500,000 workers (10 times what is shown in the figure)
and found that context switch time was around 75 ns. In comparison,
for the same yield test using kernel-level Pthreads on a single core,
the switch time is 450ns for a few threads and 800ns for 1000-32000
threads.

If we calculate aggregate context switch \emph{rate} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we empirically found that 4 cache lines (1 for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8 cache line transfers. The off-chip
bandwidth of a single socket in our system is 270M cache lines per
second \cite{porterfield:bw,Nelson:hotpar11}. This implies that
we can sustain at most 34M context switches per second per socket once
the space of all contexts is enough to mostly reside outside of L3.

In summary, our context switch engine is able to efficiently sustain very high
concurrency and as we will show later, the amount of concurrency sustained is
sufficient for the latencies \Grappa needs to hide.

\paragraph{Global memory and communication.} We measure the performance of
\Grappa's global memory and communication layers using a faithful
implementation of the giga updates per second (GUPs) benchmark, which measures cluster-wide random access bandwidth.
Read-modify-write updates are dispatched at random to a large array. This
benchmark stresses the communication layer of \Grappa separately from the
scheduler, because only a single worker is used per system node.
Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain well over a
billion updates per second with 64 nodes. Note also that when aggregation is turned off, the update rate is nearly flat. Aggregation is instrumental for good communication performance.

@@

We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput --- at the 16-node data point,
the average active tasks per core over the search is 775 and 13 for T1 and T3,
respectively. On the other hand, \Grappa performs and scales very well for T1
trees.

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{l|c|c|c|c}
&	Cora &      XMT	&        MPI	&   UPC \\ \hline
GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
BFS	& 1 &	4.33	& 14.37 & 	-- \\ 
Intsort	& 1	& 3.77 &	5.88	& -- \\
UTS (T1)	& 1	& 0.38	& -- &	0.09 \\ 
Pagerank	& 1	& 4.35	& 4.87 &	-- \\ 
\end{tabular}
\end{center}
\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
\label{tab:grappa-comparisons}
\end{table}

@@

Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
within striking distance of XMT and well beyond UPC performance (UTS only).
However, it underperforms compared to hand-tuned MPI. We believe there are two reasons for this:
First, Grappa is young; much optimization and tuning of the runtime
and the application is yet to be done. Second, the MPI implementations
we used employ algorithmic-specific optimizations---some of which
resemble of what \Grappa does automatically, like message aggregation.
Grappa's GUPS results presented earlier suggest there is significant
room for improvement.

@@

%%% I couldn't make this paragraph useful in the limited time we have.
%%% I think it's ok without. -Mark
\comment{Recall that global memory can be accessed in two ways. For requests of
locations that are expected to shared, delegate operations are
used. For requests of non-shared locations where locality can be
exploited, explicit cache operations are used. We see this reflected
in the fraction of requests that are handled remotely: less than one
percent of delegate operations in both BFS and UTS access local
data. Since explicit cache operations are intended to exploit
locality; they more often access data on the local node. Only 20
percent of UTS explicit cache operations are remote, whereas 44
percent of BFS explict cache operations are remote.}

In UTS, we obtained the best throughput by flushing frequently: on the
order of 100 microseconds. This, combined with the fewer number of remote
accesses, means that there are a limited nubmer of requests generated
per flush interval, leading to packets primarily smaller than 10KB.

In BFS, we see a trimodal distribution. There are two reasons for
this. First, in BFS, we obtained the best throughput by flushing at
least 500 microseconds apart, and also flushing whenever a core is
idle. Since this process ignores whether another flush has occurred
recently, it leads to a large number of small messages. This leads to
the first and smallest mode of the distribution.  The other two modes
are due to a locality optimization in the \Grappa version of BFS. To
minimize unnecessary remote communication, each core visits vertices
whose edge lists are stored locally. Since we are traversing power-law graphs, a small fraction of vertices have a very high number of edges. This means that occasionally a core visits a high-degree vertex which allows it to very quickly generate a large batch of requests. This interacts well with aggregation; when we find high degree vertices, we are able to aggregate packets on the order of 40KB. In flush intervals without high-degree vertices, we see packets on the order of 8KB.

@@

\comment{

@@

% Got rid of this discussion of UTS by itself, tried to work the highlights in below
%\paragraph{UTS-Mem}
%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
%nodes---the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.

%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
%scaling up through 128 nodes, although scaling
%declines because of the unscalability of our aggregation mechanism as
%number of network endpoints increases. 
%
%Despite our efforts to tune the UTS implementation specific to the 
%XMT, performance does not scale well with increasing processor count,
%flattening out around 60 processors.  When we increase the size of
%the tree from 100M to 4.2B, we find that performance does not improve,
%suggesting that performance is not limited by task parallelism.
%Cray's performance tools show an increasing number of memory
%retry operations for failed synchronization operations generated by
%the runtime, which create network contention.
%
To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes. For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance. For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count. All of the benchmarks continue to improve out to 128 nodes for \Grappa. UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return. For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa. We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.  This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).  Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.  Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.

@@

\subsection{Scaling}
% Got rid of this discussion of UTS by itself, tried to work the highlights in below
%\paragraph{UTS-Mem}
%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
%nodes---the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
%scaling up through 128 nodes, although scaling
%declines because of the unscalability of our aggregation mechanism as
%number of network endpoints increases. 
%
%Despite our efforts to tune the UTS implementation specific to the 
%XMT, performance does not scale well with increasing processor count,
%flattening out around 60 processors.  When we increase the size of
%the tree from 100M to 4.2B, we find that performance does not improve,
%suggesting that performance is not limited by task parallelism.
%Cray's performance tools show an increasing number of memory
%retry operations for failed synchronization operations generated by
%the runtime, which create network contention.
%
To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes. For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance. For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count. All of the benchmarks continue to improve out to 128 nodes for \Grappa. UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return. For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa. We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.  This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).  Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.  Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.

@@

Parallel overhead---in the form of context switches, task spawns, and
synchronization---can reduce the performance benefit of parallelism.
\Grappa sees a benefit to limiting the amount of parallelism created by
a recursive loop decomposition. The parallel loop threshold (``parallel
granularity'') parameter tells the runtime when to stop creating new tasks and just execute iterations
sequentially. This allows us to amortize the overhead of task
creation. In addition, assigning sequential iterations to a single
task provides the potential to exploit locality when data for adjacent iterations is
also adjacent in memory. The ability to exploit this locality that
exists in the application is an important advantage. We found that in UTS and BFS, increasing the
threshold from 1 up to 8 or 16, respectively, increases performance by
more than 60\%.

>>>>

The goal of our evaluation is to show that the core pieces of the \Grappa
runtime system, namely our tasking system and the global memory/communication
layer work as expected and together are able to efficiently run irregular
applications. We evaluate \Grappa in three basic steps:

@@

\item We present results that show that \Grappa can
    support large amounts of concurrency, sufficient for remote memory
    access and aggregation. The communication layer is able to
    sustain a very high rate of global memory operations. We also show the performance of a
graph kernel that stresses communication and concurrency together.
\TODO{compare to key-value stores?}

@@

Figure~\ref{fig:context-switch-exp} shows the average context switch
time as the number of workers grow. At our standard operating point
($\approx$1K workers), context switch time is on the order of 50
ns. As we add workers, the time increases slowly, but levels off: we
also ran with 500,000 workers (10~times what is shown in the figure)
and found that context switch time was around 75ns. In comparison,
for the same yield test using kernel-level Pthreads on a single core,
the switch time is 450ns for a few threads and 800ns for 1000-32000
threads.

If we calculate aggregate context switch \emph{rate\/} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not\/} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we empirically found that 4~cache lines (1~for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8~cache-line transfers. The off-chip
bandwidth of a single socket in our system is 270M cache lines per
second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).

In summary, our context switch engine is able to efficiently sustain very high
concurrency and, as we will show later, the amount of concurrency sustained is
sufficient for the latencies \Grappa needs to hide.

\paragraph{Global memory and communication.} We measure the performance of
\Grappa's global memory and communication layers using a faithful
implementation of the giga updates per second (GUPs) benchmark, which
measures cluster-wide random access bandwidth.  Read-modify-write
updates are dispatched at random to a large array. This benchmark
stresses the communication layer of \Grappa separately from the
scheduler, because only a single worker is used per system node.
Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain
well over a billion updates per second with 64~nodes. Note also that
when aggregation is turned off, the update rate is nearly
flat. Clearly aggregation is instrumental for good communication
performance.

@@

We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput -- at the 16-node data point,
the average active tasks per core over the search is 775 for T1 and 13 for T3.
On the other hand, \Grappa performs and scales very well for T1
trees.

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{l|c|c|c|c}
&	\Grappa & XMT	&        MPI	&   UPC \\ \hline
GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
BFS	& 1 &	4.33	& 14.37 & 	-- \\ 
Intsort	& 1	& 3.77 &	5.88	& -- \\
UTS (T1)	& 1	& 0.38	& -- &	0.09 \\ 
Pagerank	& 1	& 4.35	& 4.87 &	-- \\ 
\end{tabular}
\end{center}
\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
\label{tab:grappa-comparisons}
\end{table}

@@

Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
within striking distance of XMT and well beyond UPC performance (UTS only).
However, it underperforms compared to hand-tuned MPI. We believe there are two reasons for this:
First, Grappa is young; much optimization and tuning of the runtime
and the application is yet to be done. Second, the MPI implementations
we used employ algorithmic-specific optimizations -- some of which
resemble what \Grappa does automatically, like message aggregation.
Grappa's GUPS results presented earlier suggest there is significant
room for improvement.

@@

%%% I couldn't make this paragraph useful in the limited time we have.
%%% I think it's ok without. -Mark
\grappacomment{Recall that global memory can be accessed in two ways. For requests of
locations that are expected to shared, delegate operations are
used. For requests of non-shared locations where locality can be
exploited, explicit cache operations are used. We see this reflected
in the fraction of requests that are handled remotely: less than one
percent of delegate operations in both BFS and UTS access local
data. Since explicit cache operations are intended to exploit
locality; they more often access data on the local node. Only 20
percent of UTS explicit cache operations are remote, whereas 44
percent of BFS explict cache operations are remote.}

In UTS, we obtained the best throughput by flushing frequently: on the
order of 100~microseconds. This, combined with the fewer number of remote
accesses, means that there are a limited number of requests generated
per flush interval, leading to packets primarily smaller than 10KB.

In BFS, we see a trimodal distribution. There are two reasons for
this. First, in BFS, we obtained the best throughput by flushing at
least 500~microseconds apart, and also flushing whenever a core is
idle. Since this process ignores whether another flush has occurred
recently, it leads to a large number of small messages. This leads to
the first and smallest mode of the distribution.  The other two modes
are due to a locality optimization in the \Grappa version of BFS. To
minimize unnecessary remote communication, each core visits vertices
whose edge lists are stored locally. Since we are traversing power-law graphs, a small fraction of vertices have a very high number of edges. This means that occasionally a core visits a high-degree vertex which allows it to very quickly generate a large batch of requests. This interacts well with aggregation; when we find high degree vertices, we are able to aggregate packets on the order of 40KB. In flush intervals without high-degree vertices, we see packets on the order of 8KB.

@@

\grappacomment{

@@

% Got rid of this discussion of UTS by itself, tried to work the highlights in below
%\paragraph{UTS-Mem}
%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
%nodes -- the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.

%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
%scaling up through 128 nodes, although scaling
%declines because of the unscalability of our aggregation mechanism as
%number of network endpoints increases. 
%
%Despite our efforts to tune the UTS implementation specific to the 
%XMT, performance does not scale well with increasing processor count,
%flattening out around 60 processors.  When we increase the size of
%the tree from 100M to 4.2B, we find that performance does not improve,
%suggesting that performance is not limited by task parallelism.
%Cray's performance tools show an increasing number of memory
%retry operations for failed synchronization operations generated by
%the runtime, which create network contention.
%
To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes. For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance. For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count. All of the benchmarks continue to improve out to 128 nodes for \Grappa. UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return. For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa. We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.  This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).  Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.  Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.

@@

\subsection{Scaling}
% Got rid of this discussion of UTS by itself, tried to work the highlights in below
%\paragraph{UTS-Mem}
%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
%nodes -- the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
%scaling up through 128 nodes, although scaling
%declines because of the unscalability of our aggregation mechanism as
%number of network endpoints increases. 
%
%Despite our efforts to tune the UTS implementation specific to the 
%XMT, performance does not scale well with increasing processor count,
%flattening out around 60 processors.  When we increase the size of
%the tree from 100M to 4.2B, we find that performance does not improve,
%suggesting that performance is not limited by task parallelism.
%Cray's performance tools show an increasing number of memory
%retry operations for failed synchronization operations generated by
%the runtime, which create network contention.
%
To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes. For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance. For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count. All of the benchmarks continue to improve out to 128 nodes for \Grappa. UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return. For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa. We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.  This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).  Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.  Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.

@@

Parallel overhead -- in the form of context switches, task spawns, and
synchronization -- can reduce the performance benefit of parallelism.
\Grappa sees a benefit to limiting the amount of parallelism created by
a recursive loop decomposition. The parallel loop threshold (``parallel
granularity'') parameter tells the runtime when to stop creating new tasks and just execute iterations
sequentially. This allows us to amortize the overhead of task
creation. In addition, assigning sequential iterations to a single
task provides the potential to exploit locality when data for adjacent iterations is
also adjacent in memory. The ability to exploit this locality that
exists in the application is an important advantage. We found that in UTS and BFS, increasing the
threshold from 1 up to 8 or 16, respectively, increases performance by
more than 60\%.

### 2013-07-23 doc/asplos14/intro.tex
many small fixes

Irregular applications exhibit workloads, dependences, and memory accesses
that are highly sensitive to input. Classic examples of such applications
include branch and bound optimization, SPICE circuit simulation, and car crash
analysis. Important contemporary examples include processing large graphs in
the business, national security, and social network computing domains. For
these emerging applications, reasonable response time---given the sheer
amount of data---requires large multinode systems. The most broadly available
multinode systems are those built from x86 compute nodes interconnected via
Ethernet or InfiniBand. However, scalable performance of irregular
applications on these systems is elusive for two reasons:

\vspace{0.5ex}
\noindent{\bf Poor data locality and frequent communication.} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system. This results in frequent requests for small pieces of remote data. Caches are of little assistance because of low data re-use and spatial locality. Prefetching is of limited value because request locations are not known early enough. Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are ineffective because they rely on effective data partitioning and regular communication patterns. Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

\vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance. Latency of commodity networks runs anywhere from a few to hundreds of microseconds---tens of thousands of processor clock cycles.  Since irregular application tasks encounter remote references dynamically during execution and must resolve them before making further progress, stalls are frequent and lead to severely underutilized compute resources.

@@

As a general design philosophy, \Grappa trades latency for throughput.  By \emph{increasing} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.  \Grappa then exploits parallelism to overcome latency.

>>>>

Irregular applications exhibit workloads, dependences, and memory accesses
that are highly sensitive to input. Classic examples of such applications
include branch and bound optimization, SPICE circuit simulation, and car crash
analysis. Important contemporary examples include processing large graphs in
the business, national security, and social network computing domains. For
these emerging applications, reasonable response time -- given the sheer
amount of data -- requires large multinode systems. The most broadly available
multinode systems are those built from x86 compute nodes interconnected via
Ethernet or InfiniBand. However, scalable performance of irregular
applications on these systems is elusive for two reasons:

\vspace{0.5ex}
\noindent{\bf Poor data locality and frequent communication.} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system. This results in frequent requests for small pieces of remote data. Caches are of little assistance because of low temporal and spatial locality. Prefetching is of limited value because request locations are not known early enough. Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are ineffective because they rely on effective data partitioning and regular communication patterns. Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

\vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance. Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- tens of thousands of processor clock cycles.  Since irregular application tasks encounter remote references dynamically during execution and must resolve them before making further progress, stalls are frequent and lead to severely underutilized compute resources.

@@

As a general design philosophy, \Grappa trades latency for throughput.  By \emph{increasing\/} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance via work-stealing.  \Grappa then exploits parallelism to tolerate the increased latency.
